---
title: "Gene Expression Analysis in Cancer using RNA-Seq"
author: "Pranava Upparlapalli"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
    toc_depth: '2'
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
    theme: cerulean
    highlight: tango
    code_folding: hide
    fig_caption: true
    self_contained: true
    keep_md: true
    output_file: index.html
---


## ðŸ”¬ Project Overview

This project explores gene expression variation across five cancer types (BRCA, KIRC, LUAD, COAD, PRAD) using RNA-Seq count data. Differential expression was assessed using DESeq2, followed by PCA, clustering, and visualization to identify cancer-specific gene signatures.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### STEP - 1: Loading Libraries

```{r Libraries ,fig.align='center', message=FALSE, warning=FALSE, comment=" "}

library(tidyverse)
library(tidyr)
library(DESeq2)
library(ggplot2)
library(VennDiagram)
library(UpSetR)
library(RColorBrewer)
library(pheatmap)
library(reshape2)
library(pROC)
library(caret)
library(parallel)
library(mltools)
library(MASS)
library(ggfortify)
library(stats)
library(data.table)

```

### STEP - 2: Loading Data + Normalization:
```{r Data, fig.align= 'center', comment= " "}

# Set the paths for data and labels
data_path <- "data/data.csv"
labels_path <- "data/labels.csv"

data <- read.csv(data_path, header = TRUE)
labels <- read.csv(labels_path, header = TRUE)

merged_data <- merge(data, labels, by = "X", all.x = TRUE)
merged_data <- na.omit(merged_data)
merged_data$Class <- as.factor(merged_data$Class)
gc()

```

### STEP -3: Data division visualisation:
```{r Subsets, fig.align= 'center', comment= " "}

# Count samples for each cancer type
class_data <- as.data.frame(table(merged_data$Class))
colnames(class_data) <- c("Cancer_Type", "Count")
class_data$Count <- sort(class_data$Count, decreasing = TRUE)

# Create bar plot
data_division <- ggplot(class_data, aes(x = Cancer_Type, y = Count, fill = Cancer_Type)) +
  geom_bar(stat = "identity") +
  theme_gray() +
  labs(title = "Distribution of Samples Across Cancer Types",
       x = "Cancer Type",
       y = "Number of Samples")
print(data_division)
ggsave("plots/sample_distribution.png", plot = data_division, width = 8, height = 5)
gc()

```
### STEP - 4: DESeq2 Analysis
```{r, fig.align= 'center', comment= " "}

# 1. Prepare data
# This intermediate variable holds the numeric data with samples as rows
gene_data <- merged_data[, !names(merged_data) %in% c("X", "Class")]
rownames(gene_data) <- merged_data$X

# Transpose to get the final matrix: genes as rows, samples as columns
gene_expr_matrix <- t(gene_data)
gene_expr_matrix <- round(gene_expr_matrix) # Ensure integer counts

# Prepare sample information
sample_info <- data.frame(Class = merged_data$Class)
rownames(sample_info) <- merged_data$X

# 2. Create DESeq2 object
# Check for matching sample names
stopifnot(all(colnames(gene_expr_matrix) == rownames(sample_info)))
dds <- DESeqDataSetFromMatrix(countData = gene_expr_matrix,
                              colData = sample_info,
                              design = ~ Class)

# 3. Run analysis
dds <- DESeq(dds)
res <- results(dds) # Get full results for plotting

# 4. Get significant genes
res_sig <- subset(res, padj < 0.05 & !is.na(padj)) # Filter for significance
print(summary(res_sig))

sig_gene_names <- rownames(res_sig)

# Extract normalized counts for visualization
sig_genes_data <- as.data.frame(counts(dds, normalized = TRUE)[sig_gene_names, ])
sig_genes_data$Gene <- rownames(sig_genes_data)

```

```{r}
# --- MORE ROBUST Cleaning, Clustering, and Reshaping ---

# 1. Create a clean numeric matrix from sig_genes_data
gene_names <- sig_genes_data$Gene
numeric_data <- sig_genes_data[, -which(names(sig_genes_data) == "Gene")]
rownames(numeric_data) <- gene_names

# 2. Find which genes (rows) are valid for analysis
is_row_valid <- apply(numeric_data, 1, function(row) {
  all(is.finite(row)) && var(row, na.rm = TRUE) > 0
})

# 3. Create the final, clean data matrix
gene_matrix_for_clustering <- numeric_data[is_row_valid, , drop = FALSE] # drop=FALSE is safer

# 4. === ADDED SAFETY CHECK ===
# Check if we have enough genes to perform clustering
if (nrow(gene_matrix_for_clustering) >= 2) {

  # Perform clustering on the GENES
  gene_hclust <- hclust(dist(gene_matrix_for_clustering))
  ordered_gene_names <- rownames(gene_matrix_for_clustering)[gene_hclust$order]

  # Filter the original data frame to keep only the clean genes
  sig_genes_data_clean <- sig_genes_data[is_row_valid, ]

  # Reshape the CLEAN data into a long format for plotting
  sig_genes_data_long <- pivot_longer(
    sig_genes_data_clean,
    cols = -Gene,
    names_to = "Sample",
    values_to = "Expression"
  )

  # Set the Gene factor levels based on the clustering order
  sig_genes_data_long$Gene <- factor(sig_genes_data_long$Gene,
                                     levels = ordered_gene_names)
  
  print(paste("Successfully clustered", length(ordered_gene_names), "genes."))

} else {
  
  print("Skipping clustering: Fewer than 2 genes remained after filtering.")
  # Create an empty data frame so the script doesn't fail later
  sig_genes_data_long <- data.frame() 
  
}

gc()
```
### STEP - 5: DDSEQ2 - Visualisation:
```{r, fig.align= 'center', comment= " "}

# Create a volcano plot
volcano_plot <- ggplot(res, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = ifelse(padj < 0.05, "Significant", "Not Significant")), alpha = 0.6) +
  scale_color_manual(values = c("lightcoral", "gray"), guide = FALSE) +  # Remove color legend
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "blue") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "blue") +  # Added significance line
  labs(title = "Volcano Plot of Differentially Expressed Genes",
       x = "Log2 Fold Change",
       y = "-log10(Adjusted p-value)") +
  theme_gray()

print(volcano_plot)
ggsave("plots/volcano_plot.png", plot = volcano_plot, width = 8, height = 6)


# Create an MA plot
ma_plot <- ggplot(res, aes(x = baseMean, y = log2FoldChange)) +
  geom_point(aes(color = ifelse(padj < 0.05, "Significant", "Not Significant")), alpha = 0.6) +
  scale_color_manual(values = c("lightcoral", "darkgray"), guide = FALSE) +  # Remove color legend
  geom_hline(yintercept = 0, linetype = "dashed", color = "blue") +
  labs(title = "MA Plot of Differentially Expressed Genes",
       x = "Average Expression (baseMean)",
       y = "Log2 Fold Change") +
  theme_gray()

print(ma_plot)
ggsave("plots/ma_plot.png", plot = ma_plot, width = 8, height = 6)

```
### STEP - 6: PCA
```{r, fig.align= 'center', comment= " "}

# Load libraries if they aren't already
library(matrixStats)
library(cluster)
library(factoextra)

# --- PCA Step ---
# Use variance-stabilized data for best results
vst_data <- varianceStabilizingTransformation(dds, blind = FALSE)
vst_assay <- assay(vst_data)

# Filter out genes with zero variance to prevent errors
vst_assay_filtered <- vst_assay[rowVars(vst_assay) > 0, ]

# Run PCA on the samples
pca_results <- prcomp(t(vst_assay_filtered), center = TRUE, scale. = TRUE)

# Extract the PC scores for plotting
pca_scores <- as.data.frame(pca_results$x)
pca_scores$Class <- colData(dds)$Class

# Print PCA plot (colored by true cancer type)
pca_plot <- ggplot(pca_scores, aes(x = PC1, y = PC2, color = Class)) +
  geom_point(size = 1.5) +
  labs(title = "PCA of Cancer Samples", x = "PC1", y = "PC2") +
  theme_bw()
print(pca_plot)
```
### Step - 7: K- clustering
```{r, fig.align= 'center', comment= " "}
# --- K-means Clustering Step ---
# Perform K-means on the correct PCA scores
kmeans_result <- kmeans(pca_scores[, c("PC1", "PC2")], centers = 5, nstart = 25)
pca_scores$K_Cluster <- factor(kmeans_result$cluster)

# Print K-means plot (colored by algorithm-found clusters)
cluster_plot <- ggplot(pca_scores, aes(x = PC1, y = PC2, color = K_Cluster)) +
  geom_point(size = 1.5) +
  labs(title = "K-means Clustering of Cancer Samples", x = "PC1", y = "PC2", color = "Cluster") +
  theme_bw()
print(cluster_plot)
```

```{r}

# Assess cluster validity
silhouette_analysis <- silhouette(kmeans_result$cluster, dist(pca_scores[, c("PC1", "PC2")]))
print(summary(silhouette_analysis))

# Check the new cluster centroids
print(kmeans_result$centers)

# Create and plot the silhouette using factoextra
library(factoextra)
sil_plot <- fviz_silhouette(silhouette_analysis) +
  theme_minimal() +
  labs(title = "Silhouette Plot for K-means Clustering")

print(sil_plot)
ggsave("plots/silhouette_plot.png", plot = sil_plot, width = 8, height = 6)

```
```{r}
### Corrected Gene Expression Heatmap ###

# Ensure these libraries are loaded
library(pheatmap)
library(matrixStats)

# 1. Use variance-stabilized data from the previous steps
vst_data <- varianceStabilizingTransformation(dds, blind = FALSE)

# 2. Find the top 50 most variable genes to visualize
top_var_genes <- head(order(rowVars(assay(vst_data)), decreasing = TRUE), 50)
heatmap_matrix <- assay(vst_data)[top_var_genes, ]

# 3. Create an annotation bar for the sample cancer types
annotation_col <- data.frame(
  CancerType = colData(dds)$Class
)
rownames(annotation_col) <- colnames(heatmap_matrix)

# 4. Generate the heatmap and save it to a file
# 'scale = "row"' is crucial for visualizing relative gene expression
pheatmap(heatmap_matrix,
         scale = "row",
         annotation_col = annotation_col,
         show_rownames = FALSE,
         show_colnames = FALSE,
         main = "Heatmap of Top 50 Variable Genes",
         filename = "plots/heatmap_expression.png",
         width = 8,
         height = 6)
```
## âœ… Conclusion

This project demonstrates the ability of RNA-Seq analysis to uncover meaningful differences in gene expression across multiple cancer types. Techniques like DESeq2, PCA, and unsupervised clustering provided statistically and biologically relevant insights into transcriptomic profiles.

## ðŸ” Session Info

```{r}
sessionInfo()
```
